Fast inhibitory loop explains functional map development with short-range inhibition.

### 2016-12-13 paper.tex
text changes and new figures

\title{Fast inhibitory loop explains functional map development with short-range inhibition.}
\author{J\'an Antol\'ik, Andrew P. Davison}

\begin{document}
 \maketitle
\loadkeyvaluestore{/home/jan/Doc/Papers/lsv1m-paper/key_value_store.py}

@@

As pointed out by Muir et al. in their recent study \cite{Muir2014}, to solve this apparent conflict, previous topologically organized models cortical competition have either relied on the anatomically unsupported Mexican-hat profile of lateral interactions \cite{VonderMalsburg1973,CMVC}, or relied on other biologically unrealistic properties, such as selective targeting of inhibitory neurons by long range excitatory connections \cite{?} or instanteneous synaptic transmission coupled with ommission of recurrent inhibition \cite{Kang2003,Levy2011}. Furthermore, model analysis by Muir et al. \cite{Muir2014} showed that the presence of competition across the cortical surface is predicted well by the anatomy of direct excitatory and inhibitory coupling and that multisynaptic network effects are negligible. In conclusion, currently no satisfactory explanation of how topological functional organization develops in cortical networks that is consistent with the present anatomical findings exists.

All previous topologically organized models of cortical competition have either assumed instantaneous synaptic transmission \cite{Kang2003,Levy2011,VonderMalsburg1973,CMVC} or uniform transmission delays across the different intra-cortical connections \cite{?}. This is also the case of the analysis performed by Muir et al. \cite{Muir2014} which questions the ability of the accepted cortical architecture to explain cortical competition. However, a recent study by Ohana et al. \cite{Ohana2012} have revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast excitatory-to-inhibitory-to-excitatory pathway allows for the disynptic inhibition to generate the effective Mexican-hat like lateral interactions, thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.

@@

Here we describe the two models and their variants used in this study and finish with description of a measure
for assessing the extent to which a model orientation maps resamble their biological counterparts. Since this study heavily leans on methodology 
developed in our previous studies, we offer here only shortened description and refer readers to the original articles for details.

@@

In sections ? through ? we use GCAL model \cite{Stevens2013}  which is the most advanced variant of LISSOM 
(Laterally Interconnected Synergetically Self-Organizing Map) algorithm introduced by Miikkulainen et al. \cite{CMVC}, 
which itself is based on earlier Self-Organizing Map models \cite{Kohonen1982}.

@@

As a simplification, GCAL ignores the detailed temporal properties of the subcortical neural responses and of signal propagation along the various types of connections. Instead, the model ON/OFF units have a constant, sustained output, and all connections have a constant delay, independent of the physical length of that connection. The simulator operates in discrete time steps. Retinal input changes every \insval{settle_steps} time steps (and during this period is kept constant), and therefore afferent inputs to the V1 sheet(s) are effectively updated every \insval{settle_steps} steps. This process is a discrete simulation of an otherwise continuous series of changes in membrane potential due to incoming spikes and consequent generation of spikes. One such training iteration (\insval{settle_steps} steps) in the model represents one visual fixation i.e., an iteration consists of a constant retinal activation, followed by processing at the ON/OFF and cortical levels.

@@

\noindent where $\Psi_{i}(t)$ is the activation of unit $i$ taken from the set of units in the input
sheet of projection $p$ from which unit $j$ receives input (its connection field $F_{jp}$), and $\omega_{pij}$ is the connection
weight from unit $i$ in the input sheet of projection $p$ to unit $j$ in the in the output sheet of projection $p$. All connection field
weights are initialized with uniform random noise multiplied by a 2D Gaussian profile, cut off at the distance specified below.
Contributions from each projection are weighted and summed and passed via a non-linearity $f$ to calculate the activation of a cortical neuron i at time $t$:
%%
\begin{equation}
\Psi_{i}(t)= f(\sum_{p}\gamma_{p}C_{ip}(t))
\end{equation}
%%
where $\gamma_{p}$ is a constant determining the sign (negative for inhibitory) and strength of projection $p$. 
The transfer function $f$ is a half-wave rectifying function that ensures positive activation values. It has a 
variable threshold point ($\theta$) dependent on the average activity of the unit as described in the next subsection, 
but in all cases the gain is fixed at unity. Table \ref{table:param} shows the strength, initial Gaussian kernel spatial extent, and the cut-off distance values for all projections for the first form of the GCAL model examined in section \ref{sec:SM1}. Any modifications to these base parameters 
in the GCAL model variants examined in sections \ref{sec:SM2,sec:SM3} are then reported in the respective sections.

Once all \insval{settle_steps} settling steps are complete, the settled cortica activation pattern is deemed to be the response of cortical sheets to 
the presented pattern. At this point we use the response of cortical neurons to update their threshold point ($\theta$) (using the adaptation process described below) and to update the afferent weights via Hebbian learning. Cortical activity is then reset to zero, and a new pattern is presented. Note that both adaptation and learning could instead be performed at every settling step, but this would greatly decrease computational efficiency.

@@

where $\xi=\insva{homeostatic_learning_rate}$ is the time constant of the threshold adaptation,
$\mu=\insval{target_activity}$ is a constant defining the target average activity, and $\tilde{\Psi}$ is 
the recent average activity of the unit:

@@

\noindent where $\Psi(t)$ is the output of the unit at time t and $\chi=\insval{activity_average_smoothing}$ is time constant 
controling the decay of the influence of the past activities. The effect of this scaling mechanism is to bring the average activity of each 
cortical unit closer to the specified target. If the activity in a V1 unit moves away from the target during training, the threshold for activation is thus automatically raised or lowered to bring it closer to the target. Note that an alternative rule with only a single smoothing parameter (rather than
$\xi$ and  $\chi$) could be formulated, but the rule as presented here makes it simple for the modeler to set a desired target activity.

@@

In the model, as images are presented to the photoreceptors, cortical afferent connection weights $\omega_{i,j,A} from the ON/OFF sheets are adjusted once per iterations (after cortical settling is completed) using a simple Hebbian learning rule. This rule results in connections that reflect correlations between the presynaptic ON/OFF unit activities and the postsynaptic cortical response. Hebbian connection weight adjustment at each iteration is dependent on the presynaptic activity, the postsynaptic response, and the Hebbian learning rate:

@@

@@

TODO?????

@@

Model orientation maps are calculated based on the vector average method \cite{CMVC}. We first determine the prefered frequency of neurons 
across the map. Due to the simplified stereotypical stimulus used in this study (elongated Gaussian inputs) the spatial frequency 
preference of all neurons lies in a very narrow band, and we thus use the mean prefered spatial frequency across all cortical neurons as the
value for the spatial frequency parameteter across all subsequent analysis. Next, sine grating inputs that cover the full range of remaining parameter values (combinations of all orientations and phases) are presented, and for each orientation, the peak response of the neuron is recorded. The orientation preference is calculated by constructing a vector for each orientation $\theta$ (between 0 and 180\circ), with the peak response as
the length and $\theta$ as its orientation. These vectors are summed and the preferred orientation is calculated as half of the orientation of the summed vector. The selectivity is given by the magnitude of the summed vector.

@@

In order to asses whether the proposed model develops orientation maps that match the structure of those found in real 
animals we need to utilize an automatic metric that tells how close the maps are to animal data. To this end we will 
utilize a map-quality metric that we have recently developed \cite{Stevens2013} based on the empirical observation that pinwheel count 
in biological orientation maps scales linearly with hypercolumn size across many different species \cite{Kaschube2010}.
Specifically the pinwheel density per hypercolumn area (\Lambda^2) converges to \pi, when averaged across sufficently 
large cortical surface. For a detailed description of the procedure for calculating this metric we refer reader to 
our previous work \cite{Stevens2013}, but briefly, its calculation involves three steps. First, the locations of the pinwheels
in the orientation map are detrmined as the intersections of the zero contours of the real and imaginary 
components in the polar representation of the maps, thus yielding the total pinwheel count in the map. Second the 
hypercolumn size is determined as the peak in the isotropic ring-like fourier transform of the orientation maps.
Third, using these two number we can derive the pinwheel density, but to transform it to a usefull metric between 
unity (high-quality map) and zero (low quality map) we pass it through normalized gama distribution. In order to 
avoid any edge effects contaminating this measure, only the central ?x? area of the orientation maps is analyzed for
all models presented here. We have shown that this matric reliably distinguishes low and high quality maps \cite{Stevens2013}
and is a valid measure for assessing how well the model orientation maps match animal data.

@@

In this article we will proceed through multiple gradually more complex models of orientation development, addressing several 
of the major issues with modelling of this phenomenam, eventually demonstrating that the experimentally identified fast inhibitory loop is a satisfactory explanation for how short-range inhibition can support development of cortical functional organization. We will use two different computational abstractions to explore the questions at hand. In the fist part of the study we will use computational model that is derived from the LISSOM family of models \cite{CMVC}. This choice has three advantages. It allows for a very straightforward explanation of why 
fast inhibitory loop enables short-range inhibition to induce competition. It makes our explanation directly comparable to the 
extensive set of published LISSOM family models, and thus demonstrates that the solution proposed here generalizes to range of other 
functional properties. And finally the LISSOM abstraction enables very fast simulations, thus allowing us to perform 
parameter search analysis that would be otherwise computationally prohibitive. However, as we will explain further in section \ref{sec:SM4}, some simplifications made by the LISSOM abstraction, specifically the instantaneous translation of the neuronal input into it's activity, will leave 
certain questions unanswered. These will be addressed in section \ref{sec:SM4} using more detailed rate model framework.

@@

Ohana et al. \cite{Ohana2012} have shown that transmition delays between different types of cortical neurons are not uniform, specifically they found that on average the transmission delays  between excitatory neurons are $\approx$1.4\,ms, from excitatory to inhibitory neurons are $\approx$0.5\,ms, and from inhibitory to excitatory neurons are $\approx$0.98\,ms. The sample size for connections
between inhibitory cells in the study was not large enough to be quantitatively reliable. The key observation here is that the 
combined di-synaptic delay from excitatory to inhibitory and inhibitory to excitatory cells is approximately as long as the mono-synaptic delay
of the excitatory to excitatory connections. Drawing from this, the core insight of this study is that under such pattern of delays
the effective inhibitory interactions are well approximated by the convolution of the excitatory and inhibitory connection kernels, as we will
show below. This gives the effective inhibition longer range, consequently fulfilling the essential requirement for cortical competition to occur.
For didactic purposes, let us first explore one specific set of conditions under which the above statement holds exactly, we will explore the 
situation when these conditions are relaxed in subsequent sections:

@@

These conditions lead to following set of equations governing the cortico-cortical interaction.
\begin{align}
\label{eqn:first}
\begin{split}
& C_{m} \frac{dV}{dt} = \frac{(V_{rest} - V)}{R_{m}} + \frac{}{{R_{m}}}  g_{ex}(E_{ex} - V) + g_{inh}(E_{inh} - V)  \\

& R_{i}(x,t) = [\sum_{y}N_{\sigma_{e}}(\lVert y-x \rVert)R_{e}(y,t-\theta)]^+
\end{split}
\end{align}

\noindent where $R_{p}(x,t)$ is the response of neuron of type $p$ located at position $x$ at time $t$, $I_{aff}(x,t)$ is the afferent input to neuron $x$ at time $t$, $N_{\sigma}$(d) 
is normal distribution of variance $\sigma$, corresponding to the lateral connection kernel. When we expand for $R_{i}$ we obtain:

@@

\begin{equation}
R_{e}(x,t) = [\sum_{y}N_{\sigma_{e}}(\lVert y-x \rVert)R_{e}(y,t-\alpha) - \sum_{z}N_{\sqrt{2}\sigma_{e}}(\lVert z-x \rVert) R_{e}(z,t-\alpha)]^+
\end{equation}

This shows that under these specific assumptions the effective inhibitory interactions between excitatory neurons are $\sqrt{2}$ longer than excitatory and thus follow the mexican-hat like profile required for for lateral cortical competition underlying functional map development. Even though this essential condition of cortical competition is fulfilled in this model configuration, important constraints on the extent of the effective lateral
interactions (relative to lateral excitation) remain. It is thus still unclear wether development of high-quality orientation maps as observed experimentally is supported under these conditions. Conveniently, the lateral interaction in the LISSOM family of models is governed by the same equations as ? and ?. In the following we will use the GCAL model \cite{Stevens2013}, the latest and most robust addition to the LISSOM family of models to demonstrate that the above specific configuration of effective lateral excitation and inhibition permits development of high-quality orientation maps. To facilitate comparison we will use exactly the same model configuration as in Stevens et al. \cite{Stevens2013} (see Materials and Methods), except to three modifications necessitated by the analysis above:

@@

\item We will change the spread of lateral inhibition to be $\sqrt{2}$ longer than lateral excitatory spread, such that the resulting lateral
interactions conform with equation ?.

\item The change in point ? will results in change in the balance of overall excitation and inhibition in the model which is critical 
for successfull functional development in the model. We will thus modify the strength of the lateral inhibition to compensate for changes in ?.

\item Analogously to ?, the changes in ? change the overall balance between feedforward and lateral contributions to model cortical neuron's activity, 
which we will compensate by changing the overall strength of the lateral interactions.

@@

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/Figure1/figure1.png}
\caption{Orientation map development with short range inhibition and fast excitatory-to-inhibitory-to-excitatory loop. (A) Map quality (see Materials and Methods) at a range of 
lateral excitatory vs inhibitory projection strength ratios and afferent vs lateral projection strength ratios. (B-D) Functional organization in 3 example parameter configurations 
indicated by the red marks. From left to right, the orientation map, fast-fourier transform of the orientation map and afferent connection fields from the ON LGN model sheet for 25 example model V1 neurons. 
(B-C) Two examples of sub-optimal orientation maps. (D) Model configuration with the highest quality map found in this parameter search. (E) The same model configuration but in this case ran with explicit
simulation of inhibitory neurons and corresponding connections. E and D are identical confirming correctnes of our analysis.}
\label{fig:figure1}
\end{figure}

@@

Furthermore, note that the GCAL model used in figure \ref{fig:figure1} only explicitly models excitatory neurons and assumes both direct excitatory and inhibitory
interactions between them, thus corresponding to equation ?. Even though above we have shown that equation ? is equivalent to equation ? to verify 
correctness of our analysis we have ran a single simulation of the GCAL model corresponding to the parameter combination with the highest map 
quality found in figure \ref{fig:figure1}, but with explicitly simulated inhibitory population (figure \ref{fig:figure1}E). In this model we thus do not model direct inhibitory interactions between excitatory neurons, but instead add excitatory to inhibitory and inhibitory to excitatory connections of the same extend as those of excitatory to excitatory pathway (see the assumption \#5). As we have shown above this model should be mathematically equivalent to the simulations figure \ref{fig:figure1}D. Note, however, that the GCAL simulations represent a discrete approximation in both time and space of equations ? and we thus expect small numerical discrepancies. Indeed, orientation maps shown in figure \ref{fig:figure1}E  are nearly identical to those in the figure \ref{fig:figure1}D, with only minor numerical differences, confirming the validity of our approach.

@@

In the previous section we have shown that under a set of specific assumptions short-range inhibition can induce 
effective mexican-hat like interaction and thus support development of orientation maps. However not all assumptions we 
made were in line with experimental evidence. In this section we will numerically show that one of these assumptions is not necessary, 
specifically that addition of inhibitory connections does not invalidate results of the previous section.

To this end we will use the exact GCAL model configuration that we have found in previous section to posses the highest
quality orientation map (see figure \ref{fig:figure1}). We will use GCAL configuration in which we will explicitly model the inhibitory neurons
(see figure ?, and equation ?) and consequently also explicitly the excitatory to inhibitory and inhibitory to excitatory connections.
Furthermore, we will add inhibitory to inhibitory connections to the model with the same extent as that of inhibitory to excitatory
connections ($\sigma_{i} = \sigma_{e}$ in eq ?).

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/Figure2/figure2.png}
\caption{Orientation map development with short range inhibition, fast excitatory-to-inhibitory-to-excitatory loop and inhibitory to inhibitory connections. (A) Map quality (see Materials and Methods) at a range of 
inhibitory to inhibitory vs inhibitory to excitatory projection strength ratios and excitatory to inhibitory connection strengths. (B-D) Functional organization in 3 example parameter configurations 
indicated by the red marks. From left to right, the orientation map, fast-fourier transform of the orientation map and afferent connection fields from the ON LGN model sheet for 25 example model V1 neurons. }
\label{fig:figure2}
\end{figure}

>>>>

\newcommand{\etal}{\textit{et al.}}

\title{Fast inhibitory loop explains functional map development with short-range inhibition.}
\author{J\'an Antol\'ik}

\begin{document}
 \maketitle
\loadkeyvaluestore{/home/jan/Doc/Papers/fast_inh_paper/key_value_store.py}

@@

As pointed out by Muir et al. in their recent study \cite{Muir2014}, to solve this apparent conflict, previous topologically organized models cortical competition have either relied on the anatomically unsupported Mexican-hat profile of lateral interactions \cite{VonderMalsburg1973,CMVC}, or relied on other biologically unrealistic properties, such as selective targeting of inhibitory neurons by long range excitatory connections \cite{law:phd09,Rutishauser2012} or instantaneous synaptic transmission coupled with omission of recurrent inhibition \cite{Kang2003,Levy2011}. Model analysis by Muir et al. \cite{Muir2014} showed that the presence of competition across the cortical surface is predicted well by the anatomy of direct excitatory and inhibitory coupling and that multi-synaptic network effects are negligible. In conclusion, currently no satisfactory explanation of how topological functional organization dev?lops in cortical networks that is consistent with the present anatomical findings exists.

All previous topologically organized models of cortical competition have either assumed instantaneous synaptic transmission \cite{VonderMalsburg1973,Kang2003,Levy2011,Grabska-Barwinska2008}} or uniform transmission delays across the different intra-cortical connections \cite{CMVC,Antolik2011}. This is also the case of the analysis performed by Muir et al. \cite{Muir2014} which questions the ability of the accepted cortical architecture to explain cortical competition. However, a recent study by Ohana et al. \cite{Ohana2012} have revealed a specific pattern of transmission delays between different neural type combinations. Specifically they found that the excitatory to excitatory connections are slow while the excitatory to inhibitory connections are fast. In this study we explore the 
possibility that this specific transmission delay pattern is the missing link that can explain how short range inhibition can lead 
to effective cortical competition. We employ computational models to show that the fast excitatory-to-inhibitory-to-excitatory pathway allows for the disynptic inhibition to generate the effective Mexican-hat like lateral interactions, thus for the first time 
explaining competition in topologically organized cortical networks with no biologically implausible assumptions. To demonstrate the
proposed properties, in this study we focus on models of development of orientation maps, because this is the most well explored
example of competition driven functional cortical organization development, but these results generalize to development of other
cortical topological properties, as well as potentially other competition based computations.

@@

Here we describe the two models and their variants used in this study and finish with description of a measure
for assessing the extent to which a model orientation maps resemble their biological counterparts. Since this study heavily leans on methodology 
developed in our previous studies, we offer here only shortened description and refer readers to the original articles for details.

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelArchitecture/drawing.png}
\caption{General model architecture. }
\label{fig:model_architecture}
\end{figure}

@@

In sections \ref{sec:SM1} through \ref{sec:SM3} we use GCAL model \cite{Stevens2013}  which is the most advanced variant of LISSOM 
(Laterally Interconnected Synergetically Self-Organizing Map) algorithm introduced by Miikkulainen et al. \cite{CMVC}, 
which itself is based on earlier Self-Organizing Map models \cite{Kohonen1982}.

@@

As a simplification, GCAL ignores the detailed temporal properties of the sub-cortical neural responses and of signal propagation along the various types of connections. Instead, the model ON/OFF units have a constant, sustained output, and all connections have a constant delay, independent of the physical length of that connection. The simulator operates in discrete time steps. Retinal input changes every \insval{settle_steps} time steps (and during this period is kept constant), and therefore afferent inputs to the V1 sheet(s) are effectively updated every \insval{settle_steps} steps. This process is a discrete simulation of an otherwise continuous series of changes in membrane potential due to incoming spikes and consequent generation of spikes. One such training iteration (\insval{settle_steps} steps) in the model represents one visual fixation i.e., an iteration consists of a constant retinal activation, followed by processing at the ON/OFF and cortical levels.

@@

\noindent where $\Psi_{i}(t)$ is the activation of unit $i$ taken from the set of units in the input
sheet of projection $p$ from which unit $j$ receives input (its connection field $F_{jp}$), and $\omega_{pij}$ is the connection
weight from unit $i$ in the input sheet of projection $p$ to unit $j$ in the in the output sheet of projection $p$. All connection field
weights are initialized with uniform random noise multiplied by a 2D Gaussian profile, cut off at the distance specified below.
Contributions from each projection are weighted and summed and passed via a non-linearity $f$ to calculate the activation of a cortical neuron i at time $t$:
%%
\begin{equation}
\Psi_{i}(t)= f(\sum_{p}\gamma_{p}C_{ip}(t))
\end{equation}
%%
where $\gamma_{p}$ is a constant determining the sign (negative for inhibitory) and strength of projection $p$. 
The transfer function $f$ is a half-wave rectifying function that ensures positive activation values. It has a 
variable threshold point ($\theta$) dependent on the average activity of the unit as described in the next subsection, 
but in all cases the gain is fixed at unity. Table \ref{table:param} shows the strength, initial Gaussian kernel spatial extent, and the cut-off distance values for all projections for the first form of the GCAL model examined in section \ref{sec:SM1}. Any modifications to these base parameters 
in the GCAL model variants examined in sections \ref{sec:SM2} and \ref{sec:SM3} are then reported in the respective sections.

Once all \insval{settle_steps} settling steps are complete, the settled cortical activation pattern is deemed to be the response of cortical sheets to 
the presented pattern. At this point we use the response of cortical neurons to update their threshold point ($\theta$) (using the adaptation process described below) and to update the afferent weights via Hebbian learning. Cortical activity is then reset to zero, and a new pattern is presented. Note that both adaptation and learning could instead be performed at every settling step, but this would greatly decrease computational efficiency.

@@

where $\xi=\insval{homeostatic_learning_rate}$ is the time constant of the threshold adaptation,
$\mu=\insval{target_activity}$ is a constant defining the target average activity, and $\tilde{\Psi}$ is 
the recent average activity of the unit:

@@

\noindent where $\Psi(t)$ is the output of the unit at time t and $\chi=\insval{activity_average_smoothing}$ is time constant 
controlling the decay of the influence of the past activities. The effect of this scaling mechanism is to bring the average activity of each 
cortical unit closer to the specified target. If the activity in a V1 unit moves away from the target during training, the threshold for activation is thus automatically raised or lowered to bring it closer to the target. Note that an alternative rule with only a single smoothing parameter (rather than
$\xi$ and  $\chi$) could be formulated, but the rule as presented here makes it simple for the modeler to set a desired target activity.

@@

In the model, as images are presented to the photoreceptors, cortical afferent connection weights $\omega_{i,j,A}$ from the ON/OFF sheets are adjusted once per iterations (after cortical settling is completed) using a simple Hebbian learning rule. This rule results in connections that reflect correlations between the presynaptic ON/OFF unit activities and the postsynaptic cortical response. Hebbian connection weight adjustment at each iteration is dependent on the presynaptic activity, the postsynaptic response, and the Hebbian learning rate:

@@

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/FigureModelVariants/drawing.png}
\caption{The four considered model variants. }
\label{fig:model_architecture}
\end{figure}

@@

@@

Model orientation maps are calculated based on the vector average method \cite{CMVC}. We first determine the preferred frequency of neurons 
across the map. Due to the simplified stereotypical stimulus used in this study (elongated Gaussian inputs) the spatial frequency 
preference of all neurons lies in a very narrow band, and we thus use the mean preferred spatial frequency across all cortical neurons as the
value for the spatial frequency parameter across all subsequent analysis. Next, sine grating inputs that cover the full range of remaining parameter values (combinations of all orientations and phases) are presented, and for each orientation, the peak response of the neuron is recorded. The orientation preference is calculated by constructing a vector for each orientation $\theta$ (between 0 and 180$\circ$), with the peak response as
the length and $\theta$ as its orientation. These vectors are summed and the preferred orientation is calculated as half of the orientation of the summed vector. The selectivity is given by the magnitude of the summed vector.

@@

In order to asses whether the proposed model develops orientation maps that match the structure of those found in real 
animals we need to utilize an automatic metric that tells how close the maps are to animal data. To this end we will 
utilize a map-quality metric that we have recently developed \cite{Stevens2013} based on the empirical observation that pinwheel count 
in biological orientation maps scales linearly with hypercolumn size across many different species \cite{Kaschube2010}.
Specifically the pinwheel density per hypercolumn area ($\Lambda^2$) converges to $\pi$, when averaged across sufficiently 
large cortical surface. For a detailed description of the procedure for calculating this metric we refer reader to 
our previous work \cite{Stevens2013}, but briefly, its calculation involves three steps. First, the locations of the pinwheels
in the orientation map are determined as the intersections of the zero contours of the real and imaginary 
components in the polar representation of the maps, thus yielding the total pinwheel count in the map. Second the 
hypercolumn size is determined as the peak in the isotropic ring-like Fourier transform of the orientation maps.
Third, using these two number we can derive the pinwheel density, but to transform it to a useful metric between 
unity (high-quality map) and zero (low quality map) we pass it through normalized Gama distribution. 
We have shown that this metric reliably distinguishes low and high quality maps \cite{Stevens2013}
and is a valid measure for assessing how well the model orientation maps match animal data.

@@

In this article we will proceed through multiple gradually more complex models of orientation development, addressing several 
of the major issues with modeling of this phenomena, eventually demonstrating that the experimentally identified fast inhibitory loop is a satisfactory explanation for how short-range inhibition can support development of cortical functional organization. We will use two different computational abstractions to explore the questions at hand. In the fist part of the study we will use computational model that is derived from the LISSOM family of models \cite{CMVC}. This choice has three advantages. It allows for a very straightforward explanation of why 
fast inhibitory loop enables short-range inhibition to induce competition. It makes our explanation directly comparable to the 
extensive set of published LISSOM family models, and thus demonstrates that the solution proposed here generalizes to range of other 
functional properties. And finally the LISSOM abstraction enables very fast simulations, thus allowing us to perform 
parameter search analysis that would be otherwise computationally prohibitive. However, as we will explain further in section \ref{sec:SM4}, some simplifications made by the LISSOM abstraction, specifically the instantaneous translation of the neuronal input into it's activity, will leave 
certain questions unanswered. These will be addressed in section \ref{sec:SM4} using more detailed rate model framework.

@@

Ohana et al. \cite{Ohana2012} have shown that transmission delays between different types of cortical neurons are not uniform, specifically they found that on average the transmission delays  between excitatory neurons are $\approx$1.4\,ms, from excitatory to inhibitory neurons are $\approx$0.5\,ms, and from inhibitory to excitatory neurons are $\approx$0.98\,ms. The sample size for connections
between inhibitory cells in the study was not large enough to be quantitatively reliable. The key observation here is that the 
combined di-synaptic delay from excitatory to inhibitory and inhibitory to excitatory cells is approximately as long as the mono-synaptic delay
of the excitatory to excitatory connections. Drawing from this, the core insight of this study is that under such pattern of delays
the effective inhibitory interactions are well approximated by the convolution of the excitatory and inhibitory connection kernels, as we will
show below. This gives the effective inhibition longer range, consequently fulfilling the essential requirement for cortical competition to occur.
For didactic purposes, let us first explore one specific set of conditions under which the above statement holds exactly, we will explore the 
situation when these conditions are relaxed in subsequent sections:

@@

These conditions lead to following set of equations governing the cortico-cortical interaction.
\begin{align}
\label{eqn:first}
\begin{split}
& R_{e}(x,t) = [\sum_{y}N_{\sigma_{e}}(\lVert y-x \rVert)R_{e}(y,t-\theta) - N_{\sigma_{i}}(\lVert y-x \rVert)R_{i}(y,t-\theta) + I_{aff}(x,t)]^+ \\
& R_{i}(x,t) = [\sum_{y}N_{\sigma_{e}}(\lVert y-x \rVert)R_{e}(y,t-\theta)]^+
\end{split}
\end{align}

\noindent where $R_{p}(x,t)$ is the response of neuron of type $p$ located at position $x$ at time $t$, $I_{aff}(x,t)$ is the afferent input to neuron $x$ at time $t$, $N_{\sigma}$
is normal distribution of variance $\sigma$, corresponding to the lateral connection kernel. When we expand for $R_{i}$ we obtain:

@@

\begin{equation}
\label{eqn:last}
R_{e}(x,t) = [\sum_{y}N_{\sigma_{e}}(\lVert y-x \rVert)R_{e}(y,t-\alpha) - \sum_{z}N_{\sqrt{2}\sigma_{e}}(\lVert z-x \rVert) R_{e}(z,t-\alpha)]^+
\end{equation}

This shows that under these specific assumptions the effective inhibitory interactions between excitatory neurons are $\sqrt{2}$ longer than excitatory and thus follow the Mexican-hat like profile required for for lateral cortical competition underlying functional map development. Even though this essential condition of cortical competition is fulfilled in this model configuration, important constraints on the extent of the effective lateral
interactions (relative to lateral excitation) remain. It is thus still unclear whether development of high-quality orientation maps as observed experimentally is supported under these conditions. Conveniently, the lateral interaction in the LISSOM family of models is governed by the same equations as \ref{eqn:first}. In the following we will use the GCAL model \cite{Stevens2013}, the latest and most robust addition to the LISSOM family of models 
to demonstrate that the above specific configuration of effective lateral excitation and inhibition permits development of high-quality orientation maps. To facilitate comparison we will use exactly the same model configuration 
as in Stevens et al. \cite{Stevens2013} (see Materials and Methods), except to three modifications necessitated by the analysis above:

@@

\item We will change the spread of lateral inhibition to be $\sqrt{2}$ longer than lateral excitatory spread, such that the resulting lateral
interactions conform with equation \ref{eqn:last}.

\item The change in point 1 will results in change in the balance of overall excitation and inhibition in the model which is critical 
for successful functional development in the model. We will thus modify the strength of the lateral inhibition to compensate for changes due to the modification 1.

\item Analogously to 2, the changes in 1 change the overall balance between feedforward and lateral contributions to model cortical neuron's activity, 
which we will compensate by changing the overall strength of the lateral interactions.

@@

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/Figure1/figure1.png}
\caption{Orientation map development with short range inhibition and fast excitatory-to-inhibitory-to-excitatory loop. (A) Map quality (see Materials and Methods) at a range of 
lateral excitatory vs inhibitory projection strength ratios and afferent vs lateral projection strength ratios. (B-D) Functional organization in 3 example parameter configurations 
indicated by the red marks. From left to right, the orientation map, fast-Fourier transform of the orientation map and afferent connection fields from the ON LGN model sheet for 25 example model V1 neurons. 
(B-C) Two examples of sub-optimal orientation maps. (D) Model configuration with the highest quality map found in this parameter search. 
(E) The same model configuration but in this case ran with explicit simulation of inhibitory neurons and corresponding connections. E and D are nearly identical confirming correctness of our analysis.
}
\label{fig:figure1}
\end{figure}

@@

Furthermore, note that the GCAL model used in figure \ref{fig:figure1} only explicitly models excitatory neurons and assumes both direct excitatory and inhibitory
interactions between them, thus corresponding to equation \ref{eqn:last}. Even though above we have shown that equation \ref{eqn:last} is equivalent to equation \ref{eqn:first} to verify 
correctness of our analysis we have ran a single simulation of the GCAL model corresponding to the parameter combination with the highest map 
quality found in figure \ref{fig:figure1}, but with explicitly simulated inhibitory population (figure \ref{fig:figure1}E). In this model we thus do not model direct inhibitory 
interactions between excitatory neurons, but instead add excitatory to inhibitory and inhibitory to excitatory connections of the same extend as those of excitatory to excitatory pathway (see the assumption \#5). 
As we have shown above this model should be mathematically equivalent to the simulations figure \ref{fig:figure1}D. Note, however, that the GCAL simulations represent a discrete approximation 
in both time and space of equations \ref{eqn:first},\ref{eqn:last} and we thus expect small numerical discrepancies. Indeed, orientation maps shown in figure \ref{fig:figure1}E  are nearly identical to those 
in the figure \ref{fig:figure1}D, with only barely perceptible numerical differences, confirming the validity of our approach.

@@

In the previous section we have shown that under a set of specific assumptions short-range inhibition can induce 
effective Mexican-hat like interaction and thus support development of orientation maps. However not all assumptions we 
made were in line with experimental evidence. In this section we will numerically show that one of these assumptions is not necessary, 
specifically that addition of inhibitory connections does not invalidate results of the previous section.

To this end we will use the exact GCAL model configuration that we have found in previous section to posses the highest
quality orientation map (see figure \ref{fig:figure1}). We will use GCAL configuration in which we will explicitly model the inhibitory neurons
(see figure ?, and equation \ref{eqn:first}) and consequently also explicitly the excitatory to inhibitory and inhibitory to excitatory connections.
Furthermore, we will add inhibitory to inhibitory connections to the model with the same extent as that of inhibitory to excitatory
connections ($\sigma_{i} = \sigma_{e}$ in eq \ref{eqn:first}).

\begin{figure}[htpb!] 
\centering
\includegraphics[width=16cm]{./SVG/Figure2/figure2.png}
\caption{Orientation map development with short range inhibition, fast excitatory-to-inhibitory-to-excitatory loop and inhibitory to inhibitory connections. (A) Map quality (see Materials and Methods) at a range of 
inhibitory to inhibitory vs inhibitory to excitatory projection strength ratios and excitatory to inhibitory connection strengths. (B-D) Functional organization in 3 example parameter configurations 
indicated by the red marks. From left to right, the orientation map, fast-Fourier transform of the orientation map and afferent connection fields from the ON LGN model sheet for 25 example model V1 neurons. }
\label{fig:figure2}
\end{figure}

By explicitly modeling the inhibitory neurons in this model we have replaced a single parameter governing the strength of inhibitory
later interactions in the model from previous section with three new parameters that set: (1) strength of excitatory to inhibitory,
(2) inhibitory to excitatory and (3) inhibitory to inhibitory connections (all other parameters remained the same as the 
in the best parametrization found in previous section). Note that in principle there is redundancy in these parameters as
the overall strength of the projections from inhibitory neurons onto both excitatory and inhibitory populations is scaled by the 
excitatory to inhibitory projections strength. Therefore in figure \ref{fig:figure2} we have systematically varied the strength of the added inhibitory to inhibitory
projection expressed relatively to the strength of the inhibitory to inhibitory connections (which was set to 1), while also varying the 
strength of the excitatory to inhibitory projection. We have  investigated the quality of the orientation 
maps that developed under these different levels of inhibitory to inhibitory interactions. As can be seen, high quality maps 
can develop under the full range of the inhibitory to inhibitory interaction strengths, depending on the overall 
excitatory to inhibitory drive. This shows that inclusion of direct inhibitory to inhibitory interactions does not 
invalidate the results of section \ref{sec:SM3}.
